Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-06 main_new.tex
updates

We consider a lateral domain, near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes;
$\xi_a(\bx)$ denotes the temperature (in $^o C$) and $\xi_b(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$.  We let $t_a$ be a
threshold for temperature and $t_b$ a threshold for salinity, the
bivariate ES is then defined by

@@

The ES holds random indicator variables, while the associated EPs are 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$, it is easy to classify water masses; it is
challenging if the probabilities are close to $0.5$. 
The integrated Bernoulli variance (IBV) \cite{bect2019}, is defined as 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}
The measure is dominated by locations with EPs close to $0.5$.
\kc{This sentence is a bit strange; just about, you state that EPs
  close to 0.5 can be problematic since they do not categorize the
  water masses well. And here you state that there is a strong
  dominance of the average. So now you should say how the biasing can
  change to make ``V'' more effective. Otherwise this sentence is
  dangling.}

If prior knowledge provides a strong indication of cold freshwater
(for instance) in a part of the domain, it is easy to classify, and
there is likely little value in exploring those locations. Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification skill. We denote temperature measurements by
$y_{a}(\bx_i)$ and salinity measurements by $y_{b}(\bx_i)$, at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements, denoted $\by$, the conditional EPs are
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by). 
\end{equation}

Our goal is to construct designs for data gathering, that are expected
to improve the mapping and separation of water masses. When designing
AUV sampling designs, it is therefore natural to evaluate the expected
reduction in the IBV, as provided by the survey data. In this way, we
take the expectation with respect to the planned measurements $\by$;
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey. A criterion for the selection is
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed for each of the possible designs.

Recent work in literature on ESs and EPs connected to spatial
statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}), and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria in the situation with multivariate GPs.

@@

The actual calculations of the EPs and the IBV require a model
specification. Here, Gaussian modeling assumptions are used, as described next. They allow closed form evaluation of the IBVs, as derived in Section \ref{sec:sur}.

@@

We represent the bivariate random fields of sea temperature and
salinity by a Gaussian model.  At location $\bx \in \mathcal{M}$, the bivariate distribution is
\begin{equation}\label{gp_bar}
\begin{bmatrix}\xi_a(\bx) \\
\xi_b(\bx) 
\end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{a}^2(\bx) & \sigma_{a}(\bx) \sigma_b(\bx) \gamma(\bx)  \\
\sigma_{a}(\bx) \sigma_b(\bx) \gamma(\bx)  & \sigma_{b}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation means Gaussian (Normal) distributed variables specified by the mean vector and covariance matrix.
In the application below, the model parameters are specified from preliminary data: The mean varies with location, capturing the variability near the river mouth, while the variance parameters and correlation between salinity and
temperature are assumed to be constant for all locations.

For the spatial dependence among variables, 
we assume a separable correlation function with the same decay for salinity and temperature,
which is not unrealistic for a river plume. We have
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
for $i,j \in {a,b}$ and $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$ otherwise.
Moreover, we assume a stationary isotropic where the correlation function
$\rho(\bx,\bx')$ solely depends on $\bx$ and $\bx'$ through their Euclidean distance $h=\sqrt{|\bx-\bx'|^2}$.
With lots of data and prior knowledge one could possibly fit and estimate parameters of non-stationary or non-isotropic covariance function and more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}, but we do not consider that here.

For implementation on the computer we discretize the spatial domain $\mathcal{M}$ to a
set of $n$ grid locations $\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area $\Delta$. The integral expression in Eq. (\ref{measVAR}) and (\ref{sur}) are then approximated by sums over all grid cells.
This grid is identical to the possible set of design locations, as mentioned previously, defined as a waypoint graph. Using vector notation, we denote the Gaussian temperature and salinity variables on the grid and its density function by
\begin{equation}\label{prior}
    \bxi = (\xi_a(\bx_1),\xi_b(\bx_1),\ldots,\xi_a(\bx_n),\xi_b(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma) %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
Here, the length
$2 n$ vector $\bmu$ holds the expected values of temperature and
salinity at the grid locations, while the $2n \times 2n$ positive definite  matrix $\bSigma$ holds the variance-covariances of temperature and salinity variables at all grid locations.

>>>>

We consider a lateral domain, near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes;
$\xi_a(\bx)$ denotes the temperature (in $^o C$) and $\xi_b(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. \kc{might be
  clearer if you used 't' for temp and 's' for salinity. 'a' and 'b'
  are somewhat opaque especially further in the text}.  We let $t_a$ be a
threshold for temperature and $t_b$ a threshold for salinity, the
bivariate ES is then defined by:

@@

The ES holds random indicator variables, while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$, it is easy to classify water masses; it is
challenging if the probabilities are close to $0.5$. 
The integrated Bernoulli variance (IBV) \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}
The measure is dominated by locations with EPs close to $0.5$.
\kc{This sentence is a bit strange; just about, you state that EPs
  close to 0.5 can be problematic since they do not categorize the
  water masses well. And here you state that there is a strong
  dominance of the average. So now you should say how the biasing can
  change to make ``V'' more effective. Otherwise this sentence is
  dangling.}

If prior knowledge provides a strong indication of cold freshwater
(for instance) in a part of the domain, it is easy to classify, and
there is likely little value in exploring those locations. Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification skill. We denote temperature measurements by
$y_{a}(\bx_i)$ and salinity measurements by $y_{b}(\bx_i)$, at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements, denoted $\by$, the conditional EPs are:
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_a(\bx) \leq t_a, \xi_b(\bx) \leq t_b |\by). 
\end{equation}

Our goal is to construct designs for data gathering, that are expected
to improve the mapping and separation of water masses. When designing
AUV sampling designs, it is therefore natural to evaluate the expected
reduction in the IBV, as provided by the survey data. In this way, we
take the expectation with respect to the planned measurements $\by$;
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey. A criterion for the selection is:
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}), and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria in the situation with multivariate GPs.

@@

Calculations of the EPs and the IBV require a model specification.
Here, we use Gaussian modeling assumptions which allow closed form
evaluation of the IBVs, as derived in Section \ref{sec:sur}.

@@

We represent the bivariate random fields of temperature and salinity
with a Gaussian model. At location $\bx \in \mathcal{M}$, the
bivariate distribution is:
\begin{equation}\label{gp_bar}
  \begin{bmatrix}\xi_a(\bx) \\
    \xi_b(\bx) \end{bmatrix}
 \sim N \left( 
\begin{bmatrix} \mu_{a}(\bx)\\
\mu_{b}(\bx)
\end{bmatrix},\begin{bmatrix}
\sigma_{a}^2(\bx) & \sigma_{a}(\bx) \sigma_b(\bx) \gamma(\bx)  \\
\sigma_{a}(\bx) \sigma_b(\bx) \gamma(\bx)  & \sigma_{b}^2(\bx) 
\end{bmatrix}
\right),
\end{equation}
where the notation refers to Gaussian (Normal) distributed variables
specified by the mean vector and covariance matrix. In the application
below, the model parameters are specified from preliminary data: The
mean varies with location, capturing the variability near the river
mouth, while the variance parameters and correlation between salinity
and temperature are assumed to be constant for all locations. \kc{?
  Isn't that too strong an assumption? T/S tend to be correlated,
  however for uniform bodies of water; here we're dealing with coastal
  ecosystems where the two are not necessarily so.}

For the spatial dependence among variables, 
we assume a separable correlation function with the same decay for salinity and temperature,
which is not unrealistic for a river plume \kc{why so?}. We have:
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
for $i,j \in {a,b}$ and $\delta_{ij}=1$ if $i=j$ and $\delta_{ij}=0$
otherwise. Moreover, we assume a stationary isotropic property
\kc{This is probably related to the assumption as above where T/S are
  assumed to be constant in all locations? But why?} where the
correlation function $\rho(\bx,\bx')$ solely depends on $\bx$ and
$\bx'$ via a Euclidean distance measure $h=\sqrt{|\bx-\bx'|^2}$. With
data and prior knowledge one could possibly fit and estimate
parameters of non-stationary or non-isotropic covariance functions and
more complex multivariate spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however that is left for
future work.

For implementation purposes, we discretize the spatial domain
$\mathcal{M}$ to a set of $n$ grid locations
$\mathcal{M}_g = \{\bx_i, i=1,\ldots,n \}$, where each cell has area
$\Delta$. The integral expression in Eq. (\ref{measVAR}) and
(\ref{sur}) are then approximated by sums over all grid cells. This
grid is identical to the possible set of design locations and defined
as a waypoint graph \kc{You mean the
  centroids of the cells are part of the graph?}. Using vector
notation, we denote the Gaussian temperature and salinity variables on
the grid and its density function by: 
\begin{equation}\label{prior}
    \bxi = (\xi_a(\bx_1),\xi_b(\bx_1),\ldots,\xi_a(\bx_n),\xi_b(\bx_n))^t, \hspace{3mm}
    \bxi  \sim  N(\bmu, \bSigma) %\pi (\bxi) & =& \frac{1}{(2\pi)^{n}|\bSigma|^{\frac{1}{2}}}e^{-\frac{1}{2}(\bxi -\bmu)^T\bSigma^{-1}(\bxi-\bmu)}.
\end{equation}
%Here, $\pi(\bxi)$ is the joint density function of length $2 n$ vector $\bxi$, and
The length $2 n$ vector $\bmu$ holds the expected values of
temperature and salinity at the grid locations, while the
$2n \times 2n$ positive definite matrix $\bSigma$ holds the
variance-covariances of temperature and salinity variables at all grid
locations.
